From Reversible Programs to \\ Univalent Universes and Back

### 2021-05-03 Pi+/paper/univalent.tex
Add more in sec 3

Further, using the groupoid structure of $A$, we can show that transport lifts paths to equivalences, we define
$\term{transport-equiv}$.

\[
  \term{transport-equiv}(P) : \dfun{x,y:A}{x \id_{A} y \to P(x) \eqv P(y)}
\]

\begin{definition}[Univalent Fibration]
  $P$ is a univalent type family (or simply a univalent fibration) if
  $\term{transport-equiv}$ is an equivalence.
\end{definition}

@@

The univalence principle characterises paths in the universe. It says that equivalent types are equal, or the following
function is an equivalence.

@@

\begin{proof}

\end{proof}

\begin{example}[$\Aut$]
  \[
    \Aut[T] \defeq T \eqv T
  \]
\end{example}

\begin{example}[$\BAut$]
  \[
    \BAut[T] \defeq \Sub{T}
  \]
\end{example}

\begin{example}
  \[
    P(X) \defeq \SubP{X}{\Bool}
  \]
\end{example}

\begin{example}
  For any $n : \Nat$, we have
  \[
    P(X) \defeq \SubP{X}{\Fin[n]}
  \]
  \[
    \UFin[n] \defeq \Sub{\Fin[n]}
  \]
  is the univalent subuniverse of types with cardinality $n$.
\end{example}

\begin{definition}[$\isFin$]
  \[
    \isFin[X] \defeq \dsum{n:\Nat}{\SubP{X}{\Fin[n]}}
  \]
\end{definition}

@@

\begin{example}
  \[
    \UFin \defeq \dsum{X:\UU}{\isFin[X]}
  \]
  is the univalent subuniverse of all finite types.
\end{example}

@@

\begin{proposition}
  \[
    \Omega\BAut[T] \eqv \Aut[T]
  \]
\end{proposition}

>>>>

Further, using the groupoid structure of $A$, we can show that transport lifts paths to equivalences, we define

\[
  \tptEqv{P} : \dfun{x,y:A}{x \id_{A} y \to P(x) \eqv P(y)}
\]

\begin{definition}[Univalent Fibration]
  $P$ is a univalent type family (or simply a univalent fibration) if $\tptEqv{P}$ is an equivalence.
\end{definition}

@@

Voevodsky's \emph{univalence} principle characterises paths in the universe. It says that equivalent types are equal, or
the following function is an equivalence.

@@

\begin{proof}
  Suppose $(U, \El) \defeq (\dsum{X:\UU}{P(X)}, \fst)$ is a subuniverse generated by a subtype $P : \UU \to \UU$. For
  any $X, Y : \UU$ such that $\phi : P(X)$ and $\psi : P(Y)$, we want to show that
  $\tptEqv{\fst} : (X,\phi) \id (Y,\psi) \to X \eqv Y$ is an equivalence. We construct
  $X \eqv Y \to (X,\phi) \id (Y,\psi)$ by $\ua$ and using the fact that $P(\blank)$ is a proposition. That it is an
  inverse follows by calculation using the appropriate computation rules.
\end{proof}

\begin{example}[$\BAut$]
  The type of self-equivalences on a type $T$ is defined as $\Aut[T] \defeq T \eqv T$. The subuniverse of types that are
  merely equal to $T$ is given by $\BAut[T] \defeq \Sub{T}$. This turns out to be the delooping of $\Aut[T]$, as
  justified by~\cref{lem:loop-deloop}.
\end{example}

\begin{example}
  The univalent subuniverse of types merely equal to $\Bool$ is given by the subtype $P(X) \defeq \SubP{X}{\Bool}$. This
  can be used to give the denotational semantics of a 2-bit reversible
  language~\cite{caretteReversibleProgramsUnivalent2018}.
\end{example}

\begin{definition}[$\Fin$]
  The type family $\Fin : \Nat \to \UU$ is the type of finite sets indexed by cardinality. It is defined equivalently in
  two different ways,
  \begin{gather*}
    \begin{aligned}
      \Fin[n] & \defeq \dsum{k:\Nat}{k \leq n}
    \end{aligned}
    \begin{aligned}
      & \Fin[0] \defeq \bot \\
      & \Fin[\suc[n]] \defeq \top \sqcup \Fin[n]
    \end{aligned}
  \end{gather*}
  Note that $\Fin[n]$ is a $\hSet$, and we use the definitions interchangeably.
\end{definition}

\begin{example}
  For any $n : \Nat$, we define the subtype of types merely equal to $\Fin[n]$ as $P(X) \defeq \SubP{X}{\Fin[n]}$. This
  gives the univalent subuniverse $\UFin[n] \defeq \Sub{\Fin[n]}$ of finite types with cardinality equal to $n$.
\end{example}

\begin{definition}[$\isFin$]
  We say that a type is finite if it is merely equal to $\Fin[n]$ for some $n$.
  \[
    \isFin[X] \defeq \dsum{n:\Nat}{\SubP{X}{\Fin[n]}}
  \]
  Note that the natural number $n$ need not be truncated, as justified below.
\end{definition}

@@

\begin{proof}
  Suppose we have $(n,\phi) : \isFin[X]$ and $(m,\psi) : \isFin[X]$, we need to show that $(n,\phi) \id (m,\psi)$. It is
  enough to show that $n \id m$. Since $\Nat$ is a set, this is a proposition, so we can use the induction principle of
  propositional truncation to eliminate to $n \id m$, applying it on $\phi$ and $\psi$ respectively. This gives us the
  equalities $X \id \Fin[n]$ and $X \id \Fin[m]$, which gives us $\Fin[n] \id \Fin[m]$, from which $n \id m$ follows by
  taking the first projection.
\end{proof}

\begin{example}
  The univalent subuniverse of \emph{all finite types} is given by
  \[
    \UFin \defeq \dsum{X:\UU}{\isFin[X]}.
  \]
  We use this as the denotational semantics to interpret our reversible programming language $\PiPlusLang$.
\end{example}

@@

\begin{lemma}~\label{lem:loop-deloop}
  \[
    \Omega\BAut[T] \eqv \Aut[T]
  \]
\end{lemma}
